Transient Corotating Clumps Around Adolescent Low-Mass Stars From Four Years of TESS

### 2023-09-04 paper/ms.tex
update ms

%\title{Transient Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Rotators from TESS}
\title{Transient Corotating Clumps Around Adolescent Low-Mass Stars From Four Years of TESS}

@@

% PENDING DISCUSSION!
%\author[0000-0002-0531-1073]{H. Isaacson}
%\affiliation{\berkeley}
%
%\author[0000-0001-8638-0320]{A. W. Howard}
%\affiliation{\caltech}

@@

% 250 word max!
\begin{abstract}
  We study a set of pre-main-sequence stars
  with sub-solar masses, rotation periods of 0.2--2 days, optical variability amplitudes
  of a few percent, and sharp recurring light curve features that last for
  hundreds of cycles.  The sharp features are probably caused by
  eclipsing clumps of dust or gas orbiting at the corotation radius.
  We refer to this class of variable star as ``complex rotators'' (CRs).
  To improve our understanding of these
  enigmatic objects, we conducted a survey for CRs using TESS
  short-cadence data collected between 2018~July and 2022~September.
  Our search of \nstarssearched\ K and M dwarfs with $T$$<$16 and
  $d$$<$150\,pc yielded \ngoods\ high-quality CRs, most of which are
  new discoveries.  Our sample includes the brightest
  ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and oldest
  ($\approx$200\,Myr) known CRs.  One of the new discoveries, LP
  12-502, stands out for the complexity of its light curve --- with as
  many as eight flux dips per cycle --- and the detail with which the
  evolution in each dip's amplitude and phase can be tracked over
  1{,}500 cycles.  We argue that corotating gas or dust remains the
  most viable explanation for CRs. The gas or dust is probably
  entrained by the star's magnetic field, and the sharp features could
  result from a multipolar field topology, a hypothesis supported by
  correspondences between the light curves of CRs and of B stars
  that may have strong multipolar magnetic fields.
\end{abstract}

@@

CRs show broadband flux variations that can be 1-2$\times$ deeper in
the blue than in the red
\citep{2017PASJ...69L...2O,2020AJ....160...86B,2022AJ....163..144G,2023MNRAS.518.2921K}.
Dust can naturally explain this chromaticity, since it has a larger
absorption cross-section in the blue than the red
\citep[e.g.][]{1989ApJ...345..245C}.  Gas might also explain the observed
chromaticities \citep{1992oasp.book.....G}.  While bound-bound
absorption can be excluded, since it provides opacity only at narrow
resonant lines, the hydrogen opacity due to bound-free
absorption is ``jagged'' \citep[see][Figure 8.5 and
Eq.~8.8]{1992oasp.book.....G}, such that at temperatures of
$\approx$3{,}000\,K to $\approx$10{,}000\,K the opacity can be larger
at blue wavelengths than at red wavelengths.  Bound-free absorption of
${\rm H}^{-}$ is often important at such temperatures, but this
opacity source is stronger in the red than the blue, the opposite of
what is required to produce deeper dips in the blue than in the red.
Likewise, Thomson scattering is too gray to be the dominant opacity
source.

@@

We found \ngoods\ objects that showed complex
quasiperiodic behavior over at least one TESS sector.  These \ngoods\
bona fide CRs are listed in Table~\ref{tab:thetable}.  This table also
includes \nmaybes\ ambiguous CRs, whose designation is
less certain, and \ndebunked\ impostors.  We inferred ages for all but
two of the \nallcands\ objects based on memberships in young stellar assocations; we
also derived temperatures and radii using SED fitting, and 
inferred stellar masses by interpolating against stellar evolutionary
models.  We caution that our sample is far from being
volume-limited and is not even magnitude-limited:
the TESS 120-second stellar sample had a heterogeneous
selection function which may have been biased in favor of young stars
over field stars.  Previous work however has shown that $\approx$1-3\%
of M dwarfs younger than $\approx$100\,Myr show the CR phenomenon
\citep{2016AJ....152..114R,2022AJ....163..144G}.

@@

      $\approx$0.05\,$P$, match the expected transit duration of a

@@

    \item Most CR dips have durations a few times longer than $T_{\rm

@@

      the shortest dips, the clumps must also

@@

The distinction between gas or dust is important because it will
clarify whether the CR phenomenon is intrinsic, so that material
comes from the star, or extrinsic, so that it is sourced through some
generic evolutionary phase of debris disks.  This knowledge would in
turn propagate to our understanding of whether the phenomenon is
primarily teaching us about dust production and processing in gas-poor
disks, or whether it is teaching us about the ability of cold gas to
remain stable in hot stellar coronae for long durations.
Observationally, acquisition of medium- or high-resolution time-series
spectra holds a good chance at resolving the gas vs.~dust question.
Given our observed $\approx$75\% on-off duty cycles, such data must be
acquired simultaneously with photometric time-series observations
(e.g. during TESS re-observation) in order for detections and
non-detections to be interpretable.

@@

\acknowledgments
LGB is grateful for support from the Heising-Simons 51~Pegasi~b
Fellowship, and for helpful conversations with G.~Laughlin, A.~Mann,
and J.~Spake.  
This paper relied on data collected by the TESS
mission, which are available from the Mikulski Archive for
Space Telescopes (MAST).  The specific 120-second cadence observations
can be accessed via DOI\,\dataset[10.17909/t9-nmc8-f686]{https://doi.org/10.17909/t9-nmc8-f686}.  Funding for the
TESS mission is provided by NASA’s Science Mission Directorate.  TIC
402980664 in particular was observed at 120-second cadence thanks to
the TESS Guest Investigator programs G022252 (PI: J.~Schlieder;
Sectors 18, 19, 25, 26) and G04168 (PI: R.~Jayaraman; Sector 53).
LGB is also grateful for the assistance of 
S.~Yee, L.~Weiss, H.~Isaacson, and A.~Howard in
acquiring and analyzing the HIRES spectra.

>>>>

\title{Transient Corotating Clumps Around Adolescent Low-Mass Stars From Four Years of TESS}

@@

@@

% 250 word max!
\begin{abstract}
  We explore a set of pre-main-sequence stars
  with sub-solar masses, rotation periods of 0.2--2 days, optical variability amplitudes
  of a few percent, and sharp recurring light curve features that last for
  hundreds of cycles.  The sharp features are probably caused by
  eclipsing clumps of dust or gas orbiting at the corotation radius.
  We refer to this class of variable star as ``complex rotators'' (CRs).
  To improve our understanding of these
  enigmatic objects, we conducted a survey for CRs using TESS
  short-cadence data collected between 2018~July and 2022~September.
  Our search of \nstarssearched\ K and M dwarfs with $T$$<$16 and
  $d$$<$150\,pc yielded \ngoods\ high-quality CRs, most of which are
  new discoveries.  Our sample includes the brightest
  ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and oldest
  ($\approx$200\,Myr) known CRs.  One of the new discoveries, LP
  12-502, stands out for the complexity of its light curve --- with as
  many as eight flux dips per cycle --- and for the detail with which the
  evolution in each dip's amplitude and phase can be tracked over
  1{,}500 cycles.  We argue that corotating gas or dust remains the
  most viable explanation for CRs. The gas or dust is probably
  entrained by the star's magnetic field, and the sharp features could
  result from a multipolar field topology, a hypothesis supported by
  correspondences between the light curves of CRs and of B stars
  that may have multipolar magnetic fields.
\end{abstract}

@@

CRs show broadband flux variations that can be 1-2$\times$ deeper in
the blue than in the red
\citep{2017PASJ...69L...2O,2020AJ....160...86B,2022AJ....163..144G,2023MNRAS.518.2921K}.
Dust can naturally explain this chromaticity, since it has a larger
absorption cross-section in the blue than the red
\citep[e.g.][]{1989ApJ...345..245C}.  Gas might also explain the observed
chromaticities \citep{1992oasp.book.....G}.  While bound-bound
absorption can be excluded, since it provides opacity only at narrow
resonant lines, the hydrogen opacity due to bound-free
absorption is ``jagged'' \citep[see][Figure 8.5 and
Eq.~8.8]{1992oasp.book.....G}, such that at temperatures of
$\approx$3{,}000\,K to $\approx$10{,}000\,K the opacity can be larger
at blue wavelengths than at red wavelengths.  Bound-free absorption of
${\rm H}^{-}$ is often important at such temperatures, but this
opacity source is stronger in the red than the blue, the opposite of
what is required to produce deeper dips in the blue than in the red.
Likewise, Thomson scattering is too gray to be the dominant opacity
source.  From hydrogen alone, bound-free absorption therefore seems
like the most plausible opacity source.  However it remains to be
demonstrated whether a sufficient population of excited states could
be maintained, particularly given the short ($\approx$microsecond)
radiative decay timescales.

@@

We found \ngoods\ objects that showed complex
quasiperiodic behavior over at least one TESS sector.  These \ngoods\
bona fide CRs are listed in Table~\ref{tab:thetable}.  This table also
includes \nmaybes\ ambiguous CRs, whose designation is
less certain, and \ndebunked\ impostors.  We inferred ages for all but
two of the \nallcands\ objects based on memberships in young stellar assocations; we
also derived temperatures and radii using SED fitting, and 
inferred stellar masses by interpolating against stellar evolutionary
models.  We caution that our sample is far from being
volume-limited and is not even magnitude-limited:
the TESS 120-second stellar sample had a heterogeneous
selection function which may have been biased in favor of young stars
over field stars.  Previous work however has shown that $\approx$1-3\%
of M dwarfs younger than $\approx$100\,Myr show the CR phenomenon
\citep{2016AJ....152..114R,2022AJ....163..144G,2022AJ....164...80R}.

@@

      $\approx$0.05\,$P$, match the expected transit duration for a

@@

    \item Many CR dips have durations a few times longer than $T_{\rm

@@

      the shortest dips, the corotating clumps must also

@@

The distinction between gas or dust is important because it could
clarify whether the CR phenomenon is intrinsic, so that material
comes from the star, or extrinsic, so that it is sourced through some
generic evolutionary phase of debris disks.  This knowledge would in
turn propagate to our understanding of whether the phenomenon is
primarily teaching us about dust production and processing in gas-poor
disks, or whether it is teaching us about the ability of cold gas to
remain stable in hot stellar coronae for long durations.
Observationally, acquisition of medium- or high-resolution time-series
spectra holds a good chance at resolving the gas vs.~dust question.
Given our observed $\approx$75\% on-off duty cycles, such data must be
acquired simultaneously with photometric time-series observations
(e.g. during TESS re-observation) in order for detections and
non-detections to be interpretable.

@@

\acknowledgments
LGB is grateful for support from the Heising-Simons 51~Pegasi~b
Fellowship.  This paper benefited from conversations between the
authors and G.~Laughlin, A.~Mann, J.~Spake, and B.~Draine.  The paper
also relied on data collected by the TESS mission, which are available
from the Mikulski Archive for Space Telescopes (MAST).  The specific
120-second cadence observations can be accessed via
DOI\,\dataset[10.17909/t9-nmc8-f686]{https://doi.org/10.17909/t9-nmc8-f686}.
Funding for the TESS mission is provided by NASA’s Science Mission
Directorate.  TIC 402980664 in particular was observed at 120-second
cadence thanks to the TESS Guest Investigator programs G022252 (PI:
J.~Schlieder; Sectors 18, 19, 25, 26) and G04168 (PI: R.~Jayaraman;
Sector 53).  LGB is also grateful for the assistance of S.~Yee,
L.~Weiss, H.~Isaacson, and A.~Howard in acquiring and analyzing the
HIRES spectra.
